Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-21 ms.tex
Made final edits to table captions, interpretation section, conclusion section.
Addressed Johannes' comments, part 1.

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced decorated Halo Occupation Distribution (HOD) 
  model using SDSS DR7 measurements of projected galaxy clustering, $\wprp (r_{\rm p})$ 
  of galaxies in r-band luminosity-threshold samples. 
  The decorated HOD is a model for the galaxy--halo
  connection that augments the traditional HOD by allowing for the possibility of 
  {\em assembly bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides mass, $\mvir$. We demonstrate that it
  is not possible to rule out galaxy assembly bias using DR7 measurements of galaxy 
  clustering alone. Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias. These samples favor models 
  in which high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite galaxy assembly bias is significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. Assembly bias should be accounted for 
  in galaxy clustering analyses or attempts to exploit galaxy clustering to constrain cosmology. 
  In addition to presenting the first constraints on HOD models that accommodate assembly bias, our analysis 
  includes numerous improvements over previous analyses of this data set and 
  supersedes previously published results, even in the case of a standard HOD analysis.
\end{abstract}

@@

Our paper is organized as follows. In Section~\ref{section:methods},
we discuss our implementation of halo occupation models and the
parameter inference methodology. We present results
from both traditional and assembly-biased halo occupation analyses
in Section~\ref{section:results}. We summarize our results and draw
conclusions in Section~\ref{section:conclusions}.

@@

We use publicly available\footnote{\tt http://www.slac.stanford.edu/$\sim$behroozi/BPlanck\_Hlists} dark matter
halo catalogs based on the {\tt ROCKSTAR} halo-finder \citep{behroozi_rockstar11} and {\tt CONSISTENT TREES} algorithm \citep{behroozi_trees13}. In particular, we use the {\tt halotools\_alpha\_version2} version of the $z=0$ snapshot
of the {\tt `bolplanck'} catalog included with {\tt Halotools.} Halos in these catalogs are based on the virial radius
density contrast given in \citet{bryan_norman98}. When populating this catalog with mock galaxies, we only use
present-day host halos with a value of $M_{\rm peak}$ that exceeds $300$ particles. The minimum 
peak halo mass considered in our analysis has $M_{\rm peak} \ge 4.05 \times 10^5\, h^{-1}\mathrm{M}_{\odot}$ 
and this prevents us from analyzing samples with $M_r > -19$.

@@

We constrain HOD parameters based on SDSS DR7 measurements of the projected
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$,
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance
matrix of the projected correlation function available at
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2},
\end{equation}
%
(Einstein summation convention), $\Delta w_{{\rm p},i} = w_{\rm p}^{\rm
  mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the
difference between the projected two-point function predicted by the
mock catalog and the measured value in the i$^{\rm th}$ separation bin
(of 12 bins total), $C^{-1}$ is the inverse of the covariance matrix
of the measurements, and the last term in Eq.~(\ref{eq:chisquare}) is
the contribution from the difference between the predicted and
measured galaxy number density. The error on the galaxy number density
assumes Poisson statistics for both the measured and predicted galaxy
number densities.

@@

Prior to discussing our results using models that include assembly bias, we present
results of standard HOD analyses that include no treatment of assembly bias. 
In our standard HOD analyses, the parameters $\log (M_{\rm min})$,
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$ are permitted to vary. 
The results of the standard HOD analyses and all other analyses are shown in the form of 
marginalized constraints on individual parameters in Table~\ref{table:parameters}. Though the quality of 
our fits, as measured by the minimum of $\chi^2$ per Degree of Freedom (DoF), varies from case to case, 
all fits have a probability $\gtrsim 1\% $ of obtaining a higher value of $\chi^2$ by chance. 
An example of the inferred posteriors for the HOD parameters is shown in 
Figure~\ref{fig:Mr20triangle} for the sample with $\magr < -20$. We do 
not show the full posteriors for all five threshold samples in the interest of brevity.
The left-hand panels of Figure~\ref{fig:Mr19samples}, Figure~\ref{fig:Mr20samples}, and Figure~\ref{fig:Mr21samples}
show the projected correlation function data along with predictions for $\wprp(\rp)$ from 50 randomly-selected
models from the MCMC chains within $\Delta \chi^2 \le 5.89$ of the best-fit model. Under the assumption of 
a Gaussian posterior distribution, $\Delta \chi^2 = 5.89$ contains 68\% of the posterior probability for a 
five-parameter model. Note that the significant covariance in the data makes it difficult to
determine the quality of fits from visual inspection of these figures.

@@

is monotonic within the physical parameter space, we show one-sided percentiles.

@@

In addition to fitting model parameters, it is possible to ask whether or not the fits to the data warrant the use of 
additional parameters (in this case, the two parameters used to describe assembly bias). This requires a 
metric that penalizes the use of additional parameters. A commonly-used metric is the 
Bayesian Information Criterion 
\citep[BIC,][]{schwarz78}, defined as 
%
\begin{equation}
\mathrm{BIC} = -2 \mathcal{L}_{\rm max} + k\, \ln N, 
\end{equation}
%
where $\mathcal{L}_{\rm max}$ is the maximum of the likelihood, 
$k$ is the number of paramters in the model and $N$ is the number of 
data points. We compute the change in the BIC, $\Delta$BIC, by subtracting 
the BIC for the analysis with assembly bias from the BIC in the standard analysis 
in which there is no assembly bias. With this convention, $\Delta \mathrm{BIC} > 0$ 
favors assembly bias, suggesting that the data warrant the additional parametric freedom. 
Conversely, $\Delta \mathrm{BIC} < 0$ favors the standard analysis, 
suggesting that the additional freedom in the model does not improve the description of the data 
to a sufficiently great degree to support the added complexity. 
A common convention is to refer to the preference as 
{\em strong} when $\vert \Delta \mathrm{BIC}\vert \ge 5$.

>>>>

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced decorated Halo Occupation Distribution (HOD) 
  model using SDSS DR7 measurements of projected galaxy clustering, $\wprp (r_{\rm p})$ 
  of galaxies in r-band luminosity-threshold samples. 
  The decorated HOD is a model for the galaxy--halo
  connection that augments the traditional HOD by allowing for the possibility of 
  {\em galaxy assembly bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides mass, $\mvir$. We demonstrate that it
  is not possible to rule out galaxy assembly bias using DR7 measurements of galaxy 
  clustering alone. Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias. These samples favor models 
  in which high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite galaxy assembly bias is significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. Assembly bias should be accounted for 
  in galaxy clustering analyses or attempts to exploit galaxy clustering to constrain cosmology. 
  In addition to presenting the first constraints on HOD models that accommodate assembly bias, our analysis 
  includes numerous improvements over previous analyses of this data set and 
  supersedes previously published results, even in the case of a standard HOD analysis.
\end{abstract}

@@

Our paper is organized as follows. In Section~\ref{section:methods},
we discuss our implementation of halo occupation models and the
parameter inference methodology. We present results
from both traditional and assembly-biased halo occupation analyses
in Section~\ref{section:results}. We discuss and interpret our results 
in Section~\ref{section:discussion}. We summarize our results and draw
conclusions in Section~\ref{section:conclusions}.

@@

We use publicly available\footnote{\tt http://www.slac.stanford.edu/$\sim$behroozi/BPlanck\_Hlists} dark matter
halo catalogs based on the {\tt ROCKSTAR} halo-finder \citep{behroozi_rockstar11} and {\tt CONSISTENT TREES} algorithm \citep{behroozi_trees13}. In particular, we use the {\tt halotools\_alpha\_version2} version of the $z=0$ snapshot
of the {\tt `bolplanck'} catalog included with {\tt Halotools.} Halos in these catalogs are based on the virial radius
density contrast given in \citet{bryan_norman98}. When populating this catalog with mock galaxies, we only use
present-day host halos with a value of $M_{\rm peak}$ that exceeds $300$ particles. The minimum 
peak halo mass considered in our analysis has $M_{\rm peak} \ge 4.05 \times 10^{10}\, h^{-1}\mathrm{M}_{\odot}$ 
and this prevents us from analyzing samples with $M_r > -19$.

@@

We constrain HOD parameters based on SDSS DR7 measurements of the projected
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$,
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance
matrix of the projected correlation function available at
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2},
\end{equation}
%
$\Delta w_{{\rm p},i} = w_{\rm p}^{\rm mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the
difference between the projected two-point function predicted by the
mock catalog and the measured value in the i$^{\rm th}$ separation bin
(of 11 bins total), $C^{-1}$ is the inverse of the covariance matrix
of the measurements, and repeated indices are summed over. 
The last term in Eq.~(\ref{eq:chisquare}) is the contribution from the difference 
between the predicted and measured galaxy number densities. 
We consider only the eleven values of 
$w_{\rm p}(r_{\rm p})$ given by \citet{zehavi_etal11} in bins logarithmically spaced between 
$r_{\rm p}=0.17\, h^{-1}\mathrm{Mpc}$ and $r_{\rm p}=16.9\, h^{-1}\mathrm{Mpc}$. Though 
\citet{zehavi_etal11} quote values at two larger separations, we find that considering those additional 
data points adds considerably to the computational expense but alter our results insignificantly. 
The error on the galaxy number density assumes Poisson statistics for both the measured and 
predicted galaxy number densities.

@@

Prior to discussing our results using models that include assembly bias, we present
results of standard HOD analyses that include no treatment of assembly bias. 
In our standard HOD analyses, the parameters $\log (M_{\rm min})$,
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$ are permitted to vary. 
The results of the standard HOD analyses and all other analyses are shown in the form of 
marginalized constraints on individual parameters in Table~\ref{table:parameters}. Though the quality of 
our fits, as measured by the minimum of $\chi^2$ per Degree of Freedom (DoF), varies from case to case, 
all have a probability $\gtrsim 1\% $ of obtaining a higher value of $\chi^2$ by chance. 
An example of the inferred posteriors for the HOD parameters is shown in 
Figure~\ref{fig:Mr20triangle} for the sample with $\magr < -20$. We do 
not show the full posteriors for all five threshold samples in the interest of brevity.
The left-hand panels of Figure~\ref{fig:Mr19samples}, Figure~\ref{fig:Mr20samples}, and Figure~\ref{fig:Mr21samples}
show the projected correlation function data along with predictions for $\wprp(\rp)$ from 50 randomly-selected
models from the MCMC chains within $\Delta \chi^2 \le 5.89$ of the best-fit model. Under the assumption of 
a Gaussian posterior distribution, $\Delta \chi^2 = 5.89$ contains 68\% of the posterior probability for a 
five-parameter model. Note that the significant covariance in the data makes it difficult to
determine the quality of fits from visual inspection of these figures.

@@

on $A_{\rm cen,sat}$ is significant at the boundary of the permissible parameter range (e.g., $A_{\rm cen}$ for $\magr < -20$), 
we provide one-sided constraints.

@@

In addition to fitting model parameters, it is possible to ask whether or not the fits to the data warrant the use of 
additional parameters (in this case, the two parameters used to describe assembly bias). This requires a 
metric that penalizes the use of additional parameters. A commonly-used metric is the 
Bayesian Information Criterion 
\citep[BIC,][]{schwarz78}, defined as 
%
\begin{equation}
\mathrm{BIC} = -2 \mathcal{L}_{\rm max} + k\, \ln N, 
\end{equation}
%
where $\mathcal{L}_{\rm max}$ is the maximum of the likelihood, 
$k$ is the number of parameters in the model and $N$ is the number of 
data points. We compute the change in the BIC, $\Delta$BIC, by subtracting 
the BIC for the analysis with assembly bias from the BIC in the standard analysis 
in which there is no assembly bias. With this convention, $\Delta \mathrm{BIC} > 0$ 
favors assembly bias, suggesting that the data warrant the additional parametric freedom. 
Conversely, $\Delta \mathrm{BIC} < 0$ favors the standard analysis, 
suggesting that the additional freedom in the model does not improve the description of the data 
to a sufficiently great degree to support the added complexity. 
A common convention is to refer to the preference as 
{\em strong} when $\vert \Delta \mathrm{BIC}\vert \ge 5$.
